---
title: "Supplementary Information"
author: "Fredston et al."
date: 2022
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

```{r setup, echo=FALSE, results="hide",message=FALSE, warning=FALSE, include=FALSE}

# load packages
library(tidyverse)
library(here)
library(lubridate) # for standardizing date format of MHW data
library(sf)
library(rnaturalearth)
#library(kableExtra)
library(modelsummary)
library(ggpubr)
library(broom)
library(mcp)
library(lme4)
library(mgcv)
# marine heatwave data for joining with survey data
mhw_summary_sat_sst_any <- read_csv(here("processed-data","mhw_satellite_sst.csv")) # MHW summary stats from satellite SST record, using any anomaly as a MHW
mhw_summary_sat_sst_5_day <- read_csv(here("processed-data","mhw_satellite_sst_5_day_threshold.csv")) # MHW summary stats from satellite SST record, defining MHWs as events >=5 days
mhw_summary_soda_sst <-  read_csv(here("processed-data","mhw_soda_sst.csv")) # modeled SST from SODA
mhw_summary_soda_sbt <-  read_csv(here("processed-data","mhw_soda_sbt.csv")) # modeled SBT from SODA

# calendar year MHW data for plotting
# mhw_cal_yr_sat_sst_5_day <- read_csv(here("processed-data","MHW_calendar_year_satellite_sst_5_day_threshold.csv"))
# mhw_cal_yr_soda_sst <- read_csv(here("processed-data","MHW_calendar_year_soda_sst.csv"))
# mhw_cal_yr_soda_sbt <- read_csv(here("processed-data","MHW_calendar_year_soda_sbt.csv"))

# survey data 
survey_summary <-read_csv(here("processed-data","survey_biomass_with_CTI.csv"))
survey_spp_summary <- read_csv(here("processed-data","species_biomass_with_CTI.csv")) %>% 
  rename('spp'=accepted_name) %>% 
  mutate(wt_mt_log = as.numeric(wt_mt_log)) 
survey_start_times <- read_csv(here("processed-data","survey_start_times.csv"))
coords_dat <- read_csv(here("processed-data","survey_coordinates.csv"))
haul_info <- read_csv(here("processed-data","haul_info.csv"))
#biomass_time <- read_csv(here("processed-data","biomass_time.csv"))
#footprint <- read_csv(here("processed-data","spatial_standardization_summary.csv"))
med_lat <- haul_info %>% group_by(survey) %>% summarise(med_lat = median(latitude))

```

```{r, setup2, include=FALSE}
knitr::opts_chunk$set(
  message=FALSE, echo=FALSE, warning=FALSE
)
```

# Supplementary Methods

# Supplementary Tables

```{r tab-models-wt }

modeldat <- survey_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% 
  filter(!is.na(wt_mt_log), !is.na(anom_days)) %>% 
  left_join(med_lat) %>% 
  mutate(
    wt_mt_log_scale = as.numeric(scale(wt_mt_log, center=TRUE, scale=TRUE)),
    cti_log_scale =  as.numeric(scale(cti_log, center=TRUE, scale=TRUE)),
    anom_days_scale =  as.numeric(scale(anom_days, center=TRUE, scale=TRUE)),
    med_lat_scale =  as.numeric(scale(med_lat, center=TRUE, scale=TRUE)),
    depth_wt_scale =  as.numeric(scale(depth_wt, center=TRUE, scale=TRUE))
  )

# models of biomass and MHWs to compare 

# LM of MHW duration on biomass
# wt_lm <- lm(wt_mt_log_scale ~ anom_days_scale, data = modeldat)

# LMEM of MHW duration on biomass
# wt_lme <- lmer(wt_mt_log_scale ~ anom_days_scale + 1|survey, data = modeldat)

# GAM of MHW duration on biomass
# wt_gam <- gam(wt_mt_log_scale ~ s(anom_days_scale), data = modeldat )

# GAM of MHW duration on biomass with a random effect
# wt_gam_re <- gam(wt_mt_log_scale ~ s(anom_days_scale) + s(survey, bs="re"), data = modeldat  %>% mutate(survey = as.factor(survey)))

# lagged effects of MHW duration on biomass 
lags <- modeldat %>% 
  group_by(survey) %>% 
  arrange(year) %>% 
  mutate(lag1 = lag(anom_days_scale, 1),
         lag2 = lag(anom_days_scale, 2),
         lag3 = lag(anom_days_scale, 3),
         lag4 = lag(anom_days_scale, 4)) %>% 
  ungroup() %>% 
  select(wt_mt_log_scale, anom_days_scale, lag1, lag2, lag3, lag4)

mlag0 <- function(dat){
  gam( wt_mt_log_scale ~ s(as.matrix(dat)[,2:2]), data=dat ) # regular GAM
}
mlag1 <- function(dat){
  gam( wt_mt_log_scale ~ s(as.matrix(dat)[,2:3]), data=dat )
}
mlag2 <- function(dat){
  gam( wt_mt_log_scale ~ s(as.matrix(dat)[,2:4]), data=dat )
}
mlag3 <- function(dat){
  gam( wt_mt_log_scale ~ s(as.matrix(dat)[,2:5]), data=dat )
}
mlag4 <- function(dat){
  gam( wt_mt_log_scale ~ s(as.matrix(dat)[,2:6]), data=dat )
}
#wt_gam_lag1 = mlag1(lags)
#wt_gam_lag2 = mlag2(lags)
#wt_gam_lag3 = mlag3(lags)
#wt_gam_lag4 = mlag4(lags)

# breakpoint regression 
# model = list(
#   wt_mt_log_scale ~ 1 + anom_days_scale,
#   ~ 0 +anom_days_scale
# )
# wt_bp = mcp(model, data=modeldat)

# compare all biomass models 
#AIC(wt_lm, wt_lme, wt_gam, wt_gam_re, wt_gam_lag1, wt_gam_lag2, wt_gam_lag3, wt_gam_lag4)

models <- list(
  "LM" = lm(wt_mt_log_scale ~ anom_days_scale, data = modeldat),
  "LME" = lmer(wt_mt_log_scale ~ anom_days_scale + 1|survey, data = modeldat),
  "GAM" = gam(wt_mt_log_scale ~ s(anom_days_scale), data = modeldat ), 
  "GAM + RE" = gam(wt_mt_log_scale ~ s(anom_days_scale) + s(survey, bs="re"), data = modeldat  %>% mutate(survey = as.factor(survey))), 
  # "Breakpoint Regression" = mcp(list(
  # wt_mt_log_scale ~ 1 + anom_days_scale,
  # ~ 0 +anom_days_scale), data=modeldat),
  "GAM Lag 1" = mlag1(lags), 
  "GAM Lag 2" = mlag2(lags), 
  "GAM Lag 3" = mlag3(lags), 
  "GAM Lag 4" = mlag4(lags)
)

modelsummary(models) # doesn't work yet because of GAM format issues 
# lagged effects of MHW duration on biomass, by survey 
# wt_gam_lag_surv <- NULL
# for(i in unique(modeldat$survey)){
#   tmp <- modeldat %>% 
#     filter(survey==i) %>% 
#     arrange(year) %>% 
#     mutate(lag1 = lag(anom_days_scale, 1),
#            lag2 = lag(anom_days_scale, 2),
#            lag3 = lag(anom_days_scale, 3)) %>% 
#     ungroup() %>% 
#     select(wt_mt_log_scale, anom_days_scale, lag1, lag2, lag3)
#   
#   # manual stop for regions where GAM throws this error without decreasing df:
#   #  Error in smooth.construct.tp.smooth.spec(object, dk$data, dk$knots) : 
#   # A term has fewer unique covariate combinations than specified maximum degrees of freedom 
#   if(i %in% c('FR-CGFS','GOA','DFO-QCS','WCANN','PT-IBTS','NIGFS')){
#     tmp0 <- cbind(glance(  gam( wt_mt_log_scale ~ s(as.matrix(tmp)[,2:2], k=3), data=tmp )
#     ), 'survey'=i,'model'='mlag0',kreduce = TRUE)    
#     tmp1 <- cbind(glance(  gam( wt_mt_log_scale ~ s(as.matrix(tmp)[,2:3], k=3), data=tmp )
#     ), 'survey'=i,'model'='mlag1',kreduce = TRUE)    
#     tmp2 <- cbind(glance(  gam( wt_mt_log_scale ~ s(as.matrix(tmp)[,2:4], k=3), data=tmp )
#     ), 'survey'=i,'model'='mlag2',kreduce = TRUE)    
#     tmp3 <- cbind(glance(  gam( wt_mt_log_scale ~ s(as.matrix(tmp)[,2:5], k=3), data=tmp )
#     ), 'survey'=i,'model'='mlag3',kreduce = TRUE)
#   } else{
#     
#     tmp0 <- cbind(glance(mlag0(tmp)), 'survey'=i,'model'='mlag0',kreduce = FALSE)
#     tmp1 <- cbind(glance(mlag1(tmp)), 'survey'=i,'model'='mlag1',kreduce = FALSE)
#     tmp2 <- cbind(glance(mlag2(tmp)), 'survey'=i,'model'='mlag2',kreduce = FALSE)
#     tmp3 <- cbind(glance(mlag3(tmp)), 'survey'=i,'model'='mlag3',kreduce = FALSE)
#   }
#   wt_gam_lag_surv <- rbind(wt_gam_lag_surv, tmp0, tmp1, tmp2, tmp3)
# }
# ggplot(wt_gam_lag_surv) +
#   geom_col(aes(x=survey, y=AIC, color=model, group=model, fill=model), position="dodge") +
#   scale_color_brewer(palette=8, type="seq") + 
#   scale_fill_brewer(palette=8, type="seq") +
#   theme_bw()
# write_csv(wt_gam_lag_surv, here('processed-data','lag_gam_summary.csv'))
```

```{r}
# LM of MHW duration on CTI (MHW-years only)
summary(lm(cti_log_scale ~ anom_days_scale, data = modeldat %>% filter(!is.na(cti_log_scale), mhw_yes_no=="yes") ))

# LMEM of MHW duration on CTI (MHW-years only)
summary(lmer(cti_log_scale ~ anom_days_scale + 1|survey, data = modeldat %>% filter(!is.na(cti_log_scale), mhw_yes_no=="yes")))

# LM of MHW duration on depth (MHW-years only)
summary(lm(depth_wt_scale ~ anom_days_scale, data = modeldat %>% filter(mhw_yes_no=="yes")))

# LM of MHW duration + latitude on biomass 
summary(lm(wt_mt_log_scale ~ anom_days_scale + med_lat_scale + med_lat_scale * anom_days_scale, data = modeldat))
summary(lm(wt_mt_log_scale ~ anom_days_scale + med_lat_scale, data = modeldat)) 


```

# Supplementary Figures

## MHW metrics

```{r fig-mhw-metric, fig.cap="Biomass change (log ratio) and three metrics of measuring MHW impacts: severity (total anomaly in °C), duration (number of days), and cumulative mean intensity (severity divided by duration, i.e., the average anomaly over the course of the MHW)." }
gg_mhw_metric <- survey_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% 
  pivot_longer(cols=c(anom_days, anom_sev, anom_int), values_to = "value", names_to = "metric") %>% 
  mutate(metric = recode(metric, anom_days='Duration (days)',anom_sev= 'Severity (°C)',anom_int= 'Cumulative mean intensity (°C/day)')) %>% 
ggplot(aes(x=value, y=wt_mt_log)) +
  geom_point() + 
  geom_smooth(method="lm") +
  facet_wrap(~metric, scales="free_x") +
  stat_regline_equation(aes(label=..rr.label..), label.y=-2) +
  labs(x="MHW metric",y="Biomass log ratio") +
  theme_bw() + 
  NULL
gg_mhw_metric
```

```{r fig-mhw-alt, fig.cap="Biomass change (log ratio) and MHW severity from three data products. Severity is calculated as the total anomaly (°C) over the 12-month interval to which each biomass log ratio is paired. Note that the greater magnitude of severity values in OISST (see x-axis) are due to its higher (daily) temporal resolution revealing greater temperature anomalies than those that appear in the monthly SODA datasets."}
gg_mhw_alt <- mhw_summary_soda_sbt%>% 
  mutate(source = "SODA SBT") %>% 
  select(-anomMonths) %>% 
  bind_rows(mhw_summary_soda_sst %>% mutate(source = "SODA SST") %>% select(-anomMonths)) %>%
  bind_rows(mhw_summary_sat_sst_any %>% mutate(source = "OISST") %>% select(-anom_days)) %>% 
  left_join(survey_summary, by="ref_yr") %>% 
  ggplot(aes(x=anom_sev, y=wt_mt_log)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_bw() + 
  coord_cartesian(clip = "off") +
  labs(x="MHW severity (total anomaly, °C)", y="Biomass log ratio") +
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") +
  facet_wrap(~source, scales="free_x") +
  stat_regline_equation(aes(label=..rr.label..), label.y=-2) +
  NULL
gg_mhw_alt

```

# Appendix: Community change and variability

Extreme events can increase ecological variability. One possible
explanation for some of our results is that MHWs might lead to greater
increases *and* decreases in biomass. We tested for this by comparing
MHW duration to the absolute biomass log ratio and found no relationship
between the two; notably, all of the largest-magnitude biomass log
ratios ocurred in years with no MHWs or relatively short MHWs
@ref(fig:fig-abs-biomass).

```{r fig-abs-biomass, fig.cap="MHW duration and absolute value of biomass log ratio. "}

gg_mhw_biomass_point_abs <- survey_summary %>% 
  inner_join (mhw_summary_sat_sst_5_day, by="ref_yr") %>% # get MHW data matched to surveys
  ggplot(aes(x=anom_days, y=abs(wt_mt_log))) +
  geom_point() +
  theme_bw() + 
  geom_smooth(method="lm", color="blue") +
  #geom_smooth(method="lm", formula = y ~ poly(x, degree=3), color="red", se=FALSE) +
#  geom_smooth(method="loess", color="green", se=FALSE) +
  #geom_line(data = my.model.abs, aes(x=anom_days, y=wt_mt_log_abs), colour = "yellow", size=1) +
  labs(x="MHW duration (days)", y="Absolute biomass log ratio") +
    stat_regline_equation(aes(label=..rr.label..), label.x=25, label.y=2.25) +
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") 
gg_mhw_biomass_point_abs

```

# Appendix: Predictors of biomass change

We explored a number of additional predictors of biomass change in the
trawl surveys beyond marine heatwaves in the preceding 12 months (the
main text results).

### Depth

Our study explored the shifts in marine fish community composition that
may occur from species shifting their ranges into or out of the study
region in response to environmental change. However, this is not the
only spatial redistribution that may arise from warming: marine species
have been widely documented shifting deeper into the oceans (@dulvy2008
@chaikin2022 ). For this reason, we also investigated whether marine
heatwaves were associated with depth shifts of these temperate marine
fishes. We do not see a change in the depth of the fish assemblage as a
function of marine heatwave duration \@ref(fig:fig-depth-point) or
whether or not a marine heatwave occurred at all
\@ref(fig:fig-depth-hist).

```{r fig-depth-point, fig.cap="Caption"}
gg_mhw_biomass_point_depth <- survey_summary %>% inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% # get MHW data matched to surveys 
  ggplot(aes(x=anom_days, y=depth_wt_log, fill=wt_mt_log, color=wt_mt_log, group=mhw_yes_no)) + 
  geom_point(position = "jitter") + 
  geom_smooth(method="lm", color = "gray35") + 
  theme_bw() + 
  coord_cartesian(clip = "off") + 
  labs(x="Marine heatwave duration (days)", y="Depth (m, weighted by biomass)", fill="Biomass log ratio",color="Biomass log ratio") + scale_color_viridis_c() + scale_fill_viridis_c() + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") 
gg_mhw_biomass_point_depth # lm estimate is 0.3 with SE 0.16, p = 0.0575

```

```{r fig-depth-hist, fig.cap="Caption"}

gg_mhw_biomass_hist_depth <- survey_summary %>% inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% # get MHW data matched to surveys 
  mutate(mhw_yes_no = recode(mhw_yes_no, no="No Marine Heatwave", yes="Marine Heatwave")) %>% 
  ggplot(aes(x=depth_wt_log, group=mhw_yes_no, fill=mhw_yes_no, color=mhw_yes_no)) + geom_freqpoly(binwidth=0.1, alpha=0.8, size=2) + 
  scale_color_manual(values=c("#E31A1C","#1F78B4")) + 
  scale_fill_manual(values=c("#E31A1C","#1F78B4")) + 
  theme_bw() + 
  labs(x="Depth log ratio", y="Frequency") + 
  theme(legend.position = "none", legend.title = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank())

gg_mhw_biomass_hist_depth
```

### Latitude

Marine communities across latitudes have responded differently to
climate change, with some declines in species richness recorded in the
tropics and at equatorward range edges (@chaudhary2021 @hastings2020 )
and some increases in species richness recorded in colder oceans and at
poleward range edges (@hastings2020 @english ). In our dataset, the
direction or magnitude of biomass change was not related to the latitude
of the region \@ref(fig:fig-lat-point).

```{r fig-lat-point, fig.cap="Caption"}
med_lat <- haul_info %>% group_by(survey) %>% summarise(med_lat = median(latitude))

gg_mhw_biomass_point_latitude <- survey_summary %>% inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% # get MHW data matched to surveys
  left_join(med_lat)%>% 
  ggplot(aes(x=anom_days, y=wt_mt_log, fill=med_lat, color=med_lat)) +
  geom_point(position = "jitter") + 
  theme_bw() + coord_cartesian(clip = "off") + 
  labs(x="Marine heatwave duration (days)", y="Biomass log ratio", fill='Median latitude', color='Median latitude') + 
  scale_color_viridis_c() + 
  scale_fill_viridis_c() + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black")

gg_mhw_biomass_point_latitude

```

### Temporal lags

Some studies find that marine communities respond rapidly to
environmental change (e.g., @flanagan2019 ). Others suggest that
ecological responses may lag disturbances for years (e.g., @babcock2010
). We explored whether marine heatwave records from further into the
past --- specifically, 1-2 and 2-3 years before each trawl survey ---
predict biomass responses. Analogous to our findings for marine
heatwaves that occur up to 12 months before each survey reported in the
main text, we found no evidence that biomass change is associated with
marine heatwave duration from prior years \@ref(fig:fig-lag-point) .

```{r fig-lag-point, fig.cap="Caption"}

gg_mhw_biomass_point_lags <- survey_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% 
  filter(!survey %in% c('AI','GOA')) %>% 
  group_by(survey) %>% 
  arrange(year) %>% 
  mutate(anom_days_lag1 = lag(anom_days, n=1), anom_days_lag2 = lag(anom_days, n=2)) %>% select(survey, wt_mt_log, anom_days, anom_days_lag1, anom_days_lag2) %>% pivot_longer(cols=c(anom_days, anom_days_lag1, anom_days_lag2), names_to="lag", values_to="anomaly_days") %>% 
  mutate(lag = recode(lag, anom_days = "0_12",anom_days_lag1= "12_24",anom_days_lag2="24_36")) %>% 
  ggplot(aes(x=anomaly_days, y=wt_mt_log)) + 
  geom_point() + geom_smooth(method="lm") + 
  # geom_smooth(data = lm.dat1, method="lm", formula = wt_mt_log ~ anom_days, inherit.aes = FALSE, aes(x=anom_days, y=wt_mt_log)) + 
  # geom_line(data=lm.predict1, aes(x=anom_days, y=wt_mt_log), color="darkgrey", size=1, inherit.aes = FALSE) +
  theme_bw() + 
  coord_cartesian(clip = "off") + 
  labs(x="Marine heatwave duration (days)", y="Biomass log ratio") + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  facet_wrap(~lag) 

gg_mhw_biomass_point_lags
```

### Species-level predictors

Species frequently exhibit individualistic responses to global change
@fredston2021. We explored whether species taxonomy, species traits, and
regional identity help to predict species-level biomass change at all
and in the context of marine heatwaves. All fish species traits were
obtained from the database in @beukhof2019. The pattern of no
relationship between MHW duration and biomass log ratio persists when
data are grouped by trophic level \@ref(fig:fig-spp-tl), feeding mode
\@ref(fig:fig-spp-feed), or habitat \@ref(fig:fig-spp-hab).

```{r tab-lm-spp, fig.cap="C", eval=FALSE}
sppscaled <- survey_spp_summary %>% 
  left_join(mhw_summary_sat_sst_5_day) %>% 
  filter(wt_mt_log > -Inf, wt_mt_log < Inf) %>% 
  mutate(log_scaled = scale(wt_mt_log, scale=TRUE, center=TRUE),
         days_scaled = scale(anom_days, scale=TRUE, center=TRUE)) 
summary(lm(formula = "wt_mt_log ~ survey + anom_days",data=sppscaled))
lmer(formula = wt_mt_log ~ anom_days + (1|survey) + (anom_days|survey),data=sppscaled)

```

```{r fig-spp-tl, fig.cap="Biomass log ratio and MHW duration grouped by trophic level of each taxon."}
traits <- read_csv(here("raw-data","TraitCollectionFishNAtlanticNEPacificContShelf.csv")) %>% 
  select(taxon, tl, feeding.mode, habitat) %>% 
  distinct() %>% 
  group_by(taxon, feeding.mode, habitat) %>% 
  summarise(tl = mean(tl)) %>% 
  group_by(taxon) %>% 
  mutate(n=n()) 

tl <- survey_spp_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day) %>% 
  inner_join(traits %>% select(taxon, tl) %>% distinct(), by=c('spp'='taxon')) %>% 
  mutate(tl_cat = ifelse(tl>4, "4-5", ifelse(tl>3, "3-4","2-3")))

ggplot(tl %>% filter(wt_mt_log < Inf & wt_mt_log > -Inf), aes(x=anom_days, y=wt_mt_log) )+
  geom_point(size=0.5, fill="grey70", color="grey30", alpha=0.5, position = "jitter") +
  facet_wrap(~tl_cat) +
  geom_smooth(method="lm") + 
  theme_bw() + 
  labs(x="Marine heatwave duration (days)", y="Biomass log ratio") + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 

```

```{r fig-spp-feed, fig.cap="Biomass log ratio and MHW duration grouped by feeding mode of each taxon."}
feed <- survey_spp_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day) %>% 
  inner_join(traits %>% select(taxon, feeding.mode, n) %>% distinct() %>% filter(!is.na(feeding.mode)) %>% group_by(taxon) %>% mutate(n=n()) %>% filter(n==1), # get a list of taxa with exactly one feeding mode record 
             by=c('spp'='taxon')) 

ggplot(feed %>% filter(wt_mt_log < Inf & wt_mt_log > -Inf), aes(x=anom_days, y=wt_mt_log) )+
  geom_point(size=0.5, fill="grey70", color="grey30", alpha=0.5, position = "jitter") +
  facet_wrap(~feeding.mode) +
  geom_smooth(method="lm") + 
  theme_bw() + 
  labs(x="Marine heatwave duration (days)", y="Biomass log ratio") + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 

```

```{r fig-spp-hab, fig.cap="Biomass log ratio and MHW duration grouped by habitat preference of each taxon."}
hab <- survey_spp_summary %>% 
  inner_join(mhw_summary_sat_sst_5_day) %>% 
  inner_join(traits %>% select(taxon, habitat, n) %>% distinct() %>% filter(!is.na(habitat)) %>% group_by(taxon) %>% mutate(n=n()) %>% filter(n==1), # get a list of taxa with exactly one feeding mode record 
             by=c('spp'='taxon')) 

ggplot(hab %>% filter(wt_mt_log < Inf & wt_mt_log > -Inf), aes(x=anom_days, y=wt_mt_log) )+
  geom_point(size=0.5, fill="grey70", color="grey30", alpha=0.5, position = "jitter") +
  facet_wrap(~habitat) +
  geom_smooth(method="lm") + 
  theme_bw() + 
  labs(x="Marine heatwave duration (days)", y="Biomass log ratio") + 
  geom_hline(aes(yintercept=0), linetype="dashed", color="black") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) 

```

Of the `r length(unique(survey_spp_summary$spp))` taxa used in the
analysis, `r length(unique(tl$spp))` had trophic level data (used in
\@ref(ref:fig-spp-tl)), `r length(unique(feed$spp))` had feeding mode
data (used in \@ref(ref:fig-spp-feed)), and `r length(unique(hab$spp))`
had feeding mode data (used in \@ref(ref:fig-spp-hab)).

# Appendix: Case study of the Northeast Pacific

The Northeast Pacific 2014-2016 is perhaps the best-documented and most
intense MHW in modern history.

```{r fig-ebs-spp, fig.cap="Biomass trends and historical MHWs in the top species (measured by biomass) in the NE Pacific. In each region, a small number of species is responsible for most of the biomass (the sixth most common species contributes <20% of the biomass as the most common species, for example)."}
topspp <- survey_spp_summary %>% 
  filter(survey %in% c('EBS','GOA','AI','WCANN')) %>% 
  group_by(survey, spp) %>% 
  summarise(tot = sum(wt_mt)) %>% 
  arrange(-tot) %>% 
  slice(1:5) %>% 
  mutate(id = paste0(survey, spp))

cbpal1 <- c("#9F0162","#009F81","#FF5AAF","#00FCCF","#8400CD","#008DF9","gray50","#00C2F9","#FFB2FD","#A40122","#E20134","#FFC33B") # http://mkweb.bcgsc.ca/colorblind/palettes.mhtml#page-container

cbpal2 <- c('#68023F','#008169','#EF0096','#00DCB5','#FFCFE2','#003C86','#9400E6','#009FFA','#FF71FD','grey50','#7CFFFA','#6A0213','#008607','#F60239','#00E307','#FFDC3D')

boop <- survey_spp_summary %>% 
  mutate(id = paste0(survey, spp)) %>% 
  filter(id %in% topspp$id) %>% 
  bind_rows(survey_spp_summary %>% filter(survey %in% c('EBS','GOA','AI'), !spp %in% topspp$spp) %>% group_by(ref_yr, year, survey) %>% summarise(wt_mt = sum(wt_mt)) %>% mutate(spp="Others")) %>% # get "others" group 
  filter(survey=="AI") %>% 
  inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% 
  mutate(bar1 = ifelse(mhw_yes_no == "yes", max(wt_mt, na.rm=TRUE), 0)) %>% 
  ggplot() +
  geom_area(aes(x=year, y=wt_mt, color=spp, fill=spp, group=spp)) + 
  geom_col(aes(x=year, y=bar1), color="transparent", fill="grey70", alpha=0.5) +
  scale_y_continuous(limits=c(0,36), breaks=seq(0, 35, 5)) +
  scale_color_manual(values=cbpal2) +
  scale_fill_manual(values=cbpal2) +
  #  facet_wrap(~survey, scales = "free") +
  theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        legend.position="bottom") + 
  labs(x="Year",y="Biomass (mt)") +
  NULL

# survey_spp_summary %>% 
#   inner_join(mhw_summary_sat_sst_5_day, by="ref_yr") %>% 
#     filter(survey %in% c('EBS','GOA','AI'),
#            spp %in% topspp$spp,
#          wt_mt_log > -Inf, 
#          wt_mt_log < Inf) %>% 
#   bind_rows(survey_summary %>% filter(survey %in% c('EBS','GOA','AI')) %>% select(ref_yr, year, survey, wt_mt_log) %>% distinct() %>% mutate(spp="All") %>% filter(!is.na(wt_mt_log))) %>% 
#   mutate(bar1 = ifelse(mhw_yes_no == "yes", max(wt_mt_log), 0), 
#          bar2 = ifelse(mhw_yes_no == "yes", min(wt_mt_log), 0)) %>% 
#   ggplot() +
#   geom_col(aes(x=year, y=bar1), color="grey70", line = "white") +
#   geom_col(aes(x=year, y=bar2), color="grey70", line = "white") +
#   geom_line(aes(x=year, y=wt_mt_log, color=spp)) + 
#   scale_y_continuous(limits=c(-5.5, 5.5), breaks=seq(-5, 5, 1)) +
#   scale_color_manual(values=cbpal) +
#   facet_wrap(~survey) +
#   theme_bw() + 
#     theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
#           legend.position="bottom") + 
#   labs(x="Year",y="Biomass log ratio") +
# NULL

```

# References
