---
title: "Exploring MHW effects on fish abundance in trawl surveys"
author: "Alexa Fredston"
date: "11/20/2020"
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
---

# Exploring marine heatwaves 

## Marine heatwave info

### From Thomas Frölicher:

The SST is obtained from the NOAA/NCEI 1/4 degree daily optimum interpolation sea surface temperature (OISST) Analysis, Version 2.1 (Banzon et al. 2020, Reynolds et al. 2007). 

Marine heatwaves (extreme SST anomalies) are defined with respect to a seasonally varying 95 percentile after detrending the SST dataset with a linear trend. The mean seasonal cycle has been additionally smoothed using a 30-day running mean to remove noise on a daily scale associated with the relative short SST record (similar approach as used in Le Grix et al.: https://bg.copernicus.org/preprints/bg-2020-412/). I averaged the SST over the individual survey areas.  Here, in contrast to William’s paper, I linearly detrended the data, whereas William used a 7-year running mean. I don’t think that this heavily influences our results. Please see below in more detail the different steps applied to the SST data:

1. delete leap days
2. calculate climatological seasonal cycle 
3. smooth climatological seasonal cycle with 30-day running mean
4. calculate SST anomalies relative to smoothed seasonal cycle 
5. detrend SST anomalies with linear trend
6. calculate SST averages over survey area
7. calculate 95P for each survey area
8. calculate anomalies to 95P for each survey area

## North America

```{r include=FALSE}
knitr::opts_chunk$set(eval=TRUE, echo=FALSE, results=TRUE, message=FALSE, warning=FALSE) # set options for all chunks
```

```{r packages, results=FALSE}
library(tidyverse)
library(here)
library(lubridate) # for standardizing date format of MHW data
```

```{r MHWdata, results=FALSE}
mhwRaw <- read_delim(here("processed-data","MHW_95P_surveys_for_malin.csv"), ";") # file from Thomas - download and move to processed-data folder

mhwYrRaw <- read_delim(here("processed-data","mhws_90P_surveys_yearly.csv"), ";") 

mhwTidy <- mhwRaw %>% 
  rename("dateRaw"=X1) %>% 
  pivot_longer(cols=2:ncol(mhwRaw), names_to="region", values_to="sstAnom") %>% # convert to "long" format
  mutate(date = dmy(dateRaw)) %>% # standardize date formats
  select(-dateRaw) %>% 
  mutate(year = year(date)) %>% 
  group_by(region, year) %>% 
  mutate(anomDays = sum(sstAnom>0, na.rm=TRUE),# count number of non-NA anomaly days 
         anomSev = sum(sstAnom, na.rm=TRUE), # add up total anomaly values
         anomIntC = anomSev / anomDays # calculate cumulative mean intensity for every region*year 
  )  %>% 
  ungroup()

mhwYrTidyPrep <- mhwYrRaw %>% 
  rename("year"=X1) %>% 
  pivot_longer(cols=2:ncol(mhwRaw), names_to="region", values_to="sstYrAnom") # convert to "long" format

mhwYrTidy <- mhwTidy %>% 
  select(region, year, anomDays, anomIntC, anomSev) %>% 
  distinct() %>% 
  mutate(anomIntC = replace_na(anomIntC, 0)) %>%  # replace NA values from calculating anomIntC with 0s
  left_join(mhwYrTidyPrep) # add in yearly MHW data 

namerRegs <- c("aleutian_islands","eastern_bering_sea","gulf_of_alaska", "gulf_of_mexico","northeast","scotian_shelf","southeast","west_coast")
```

```{r trawldata, results=FALSE}
namer_abund <- read_csv(here("processed-data","biomass_time_namer.csv"))

# stats pooled across all species 
namerSummary <- namer_abund %>% 
  filter(year>=1982) %>% # start of MHW data
  mutate(region = gsub(" ","_",str_to_lower(region)), # reformat to be consistent with MHW region IDs
         region = gsub("_us","", region)) %>% 
  group_by(region, year) %>% 
  summarise(wt = sum(sumwt)) %>% # get total weight across all species for the region*year
  mutate(wtMt = wt/1000) %>% # convert to metric tons from kg
  group_by(region) %>% 
  arrange(year) %>% 
  mutate(wtMtDiff = wtMt - lag(wtMt), # calculate first difference (delta from last year surveyed--not always a one-year lag though!)
         wtMtDiffProp = (wtMtDiff / lag(wtMt)), # calculate proportional change
         wtMtAnom = wtMt - mean(wtMt), # subtract mean over all years within the region
         wtMtAnomProp = (wtMtAnom / mean(wtMt)) # get proportional difference of anomaly relative to mean
  ) 

# species-level stats
namerSppSummary <- namer_abund %>% 
  filter(year>=1982) %>% 
  mutate(region = gsub(" ","_",str_to_lower(region)),
         region = gsub("_us","", region)) %>% 
  mutate(wtMt = sumwt/1000) %>% 
  select(-sumwt) %>% 
  group_by(region, spp) %>% 
  arrange(year) %>% 
  mutate(wtMtDiff = wtMt - lag(wtMt), # calculate first difference (delta from last year; will be 3 years ago in WCTRI)
         wtMtDiffProp = (wtMtDiff / lag(wtMt)), # calculate proportional change
         wtMtDiffPropAbs = abs(wtMtDiffProp), 
         wtMtDiffPropAbsLog = log(wtMtDiffPropAbs),
         wtMtAnom = wtMt - mean(wtMt), # subtract mean over all years within the region
         wtMtAnomProp = (wtMtAnom / mean(wtMt)) # get proportional difference of anomaly relative to mean
  ) 


# get highest-biomass spp in each region 
topspp <- namerSppSummary %>% 
  group_by(region, spp) %>% 
  summarise(wtTotal = sum(wtMt)) %>% 
  group_by(region) %>% 
  arrange(-wtTotal) %>% 
  slice_head(n=5) %>%  # change this for different no. spp/reg
  mutate(sppReg = paste0(region, spp))

```

### Newer plots

Let's plot several marine heatwave indices against several CPUE indices, to see how similar patterns are across metrics. I'm also looking for whether we see effects of major known MHW events: 2012 in the Northeast, the 2014-2015 "warm blob" on the West Coast, and the 2014-2016 heatwave and partially sea ice-free period in the Eastern Bering Sea. (The other such event in the Eastern Bering Sea was 2001-2005, which also appears to show up here.)

In addition to summing all CPUE (which will be driven by common species) I've calculated proportional change first at the single-species level and then averaged across all species in a region. This gives all species equal weight while also standardizing and centering catch values. This is shown below both as an absolute value (magnitude of change only) and from averaging across true units of single-species values. 


```{r namer-mhw-cpue-metrics, fig.cap="MHW Metrics vs. Sum CPUE Change", fig.height=7}

sppdat <- namerSppSummary %>% 
  group_by(region) %>% 
  filter(year > min(year)) %>% 
  group_by(region, year) %>% 
  summarise(wtMtAnomPropMean = mean(wtMtAnomProp, na.rm=TRUE),
            wtMtAnomPropMeanAbs = mean(abs(wtMtAnomProp), na.rm=TRUE)) 

namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982) %>% 
  left_join(sppdat) %>% 
  ggplot(aes(x=year)) + 
  geom_line(aes(y=anomIntC, color="Cumulative Mean MHW Intensity")) +
  geom_line(aes(y=anomDays/100, color="Number of SST Anomaly-Days Divided by 100")) +
  geom_line(aes(y=anomSev/100, color="MHW Severity (year sum) Divided by 100")) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  geom_line(aes(y=wtMtAnomPropMean, color="Mean Single-Species CPUE Z-Score")) +
  geom_line(aes(y=wtMtAnomPropMeanAbs, color="Mean Absolute Single-Species CPUE Z-Score")) +
  geom_point(aes(y=sstYrAnom, color="Yearly MHW Index"), shape=8, size=2.5) +
  scale_color_manual(name = "", values = c("Cumulative Mean MHW Intensity" = "darkblue",
                                           "Number of SST Anomaly-Days Divided by 100" = "turquoise1",
                                           "MHW Severity (year sum) Divided by 100" = "royalblue1",
                                           "Yearly MHW Index" = "darkblue",
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange",
                                           
                                           "Mean Single-Species CPUE Z-Score" = "palevioletred2",
                                           "Mean Absolute Single-Species CPUE Z-Score" = "mediumvioletred")) +
  facet_wrap(~region, ncol=3) +
  labs(x="Year") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  guides(color=guide_legend(nrow=4,byrow=TRUE)) +
  NULL

ggsummary <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982) %>% 
  left_join(sppdat) %>% 
  ggplot(aes(x=year)) + 
  geom_line(aes(y=anomIntC, color="Cumulative Mean MHW Intensity")) +
  geom_line(aes(y=anomDays/100, color="Number of SST Anomaly-Days Divided by 100")) +
  geom_line(aes(y=anomSev/100, color="MHW Severity (year sum) Divided by 100")) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  geom_line(aes(y=wtMtAnomPropMean, color="Mean Single-Species CPUE Z-Score")) +
  geom_line(aes(y=wtMtAnomPropMeanAbs, color="Mean Absolute Single-Species CPUE Z-Score")) +
  geom_point(aes(y=sstYrAnom, color="Yearly MHW Index"), shape=8, size=2.5) +
  scale_color_manual(name = "", values = c("Cumulative Mean MHW Intensity" = "darkblue",
                                           "Number of SST Anomaly-Days Divided by 100" = "turquoise1",
                                           "MHW Severity (year sum) Divided by 100" = "royalblue1",
                                           "Yearly MHW Index" = "darkblue",
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange",
                                           
                                           "Mean Single-Species CPUE Z-Score" = "palevioletred2",
                                           "Mean Absolute Single-Species CPUE Z-Score" = "mediumvioletred")) +
  facet_wrap(~region, ncol=4) +
  labs(x="Year") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  guides(color=guide_legend(nrow=4,byrow=TRUE)) +
  NULL

ggsave(ggsummary, filename=here("results","line_plots_all_CPUE_MHW.png"), dpi=300, width=8, height=5)
```

```{r namer-mhw-years-cpue-scatter, fig.cap="Biomass Change vs MHW-Years", fig.height=7}

namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982, !is.na(wtMtDiffProp)) %>% 
  mutate(mhwYesNo = ifelse(is.na(sstYrAnom), "No","Yes")) %>% 
  ggplot(aes(x=year, y=wtMtDiffProp, color=mhwYesNo, fill=mhwYesNo)) +
  scale_color_manual(values=c("black","red")) + 
    scale_fill_manual(values=c("black","red")) + 
  geom_point(size=3) +
  theme_bw() + 
  labs(x="Year",y="Prop. Change CPUE vs. Last Year")

ggscatter <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982, !is.na(wtMtDiffProp)) %>% 
  mutate(mhwYesNo = ifelse(is.na(sstYrAnom), "No","Yes")) %>% 
  ggplot(aes(x=year, y=wtMtDiffProp, color=mhwYesNo, fill=mhwYesNo)) +
  scale_color_manual(values=c("black","red")) + 
    scale_fill_manual(values=c("black","red")) + 
  geom_point(size=3) +
  theme_bw() + 
  labs(x="Year",y="Prop. Change CPUE vs. Last Year")

ggsave(ggscatter, filename=here("results","scatter_plots_all_CPUE_MHW.png"), dpi=300, width=8, height=5)

```


```{r namer-mhw-years-cpue-hist, fig.cap="Biomass Change vs MHW-Years", fig.height=7}

namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982, !is.na(wtMtDiffProp)) %>% 
  mutate(mhwYesNo = ifelse(is.na(sstYrAnom), "No","Yes")) %>% 
  ggplot(aes(x=wtMtDiffProp, group=mhwYesNo, color=mhwYesNo, fill=mhwYesNo)) +
  scale_color_manual(values=c("black","red")) + 
    scale_fill_manual(values=c("black","red")) + 
  geom_histogram() +
  theme_bw()

gghist <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  filter(year>=1982, !is.na(wtMtDiffProp)) %>% 
  mutate(mhwYesNo = ifelse(is.na(sstYrAnom), "No","Yes")) %>% 
  ggplot(aes(x=wtMtDiffProp, group=mhwYesNo, color=mhwYesNo, fill=mhwYesNo)) +
  scale_color_manual(values=c("black","red")) + 
    scale_fill_manual(values=c("black","red")) + 
  geom_histogram() +
  theme_bw()

ggsave(gghist, filename=here("results","histograms_all_CPUE_MHW.png"), dpi=300, width=8, height=5)

```



# Older plots / models

### Characterizing MHWs in North America

How frequent and severe have heatwaves been in North America since 1982 (the start of the MHW data)? 

```{r namer-mhw-freq,  fig.cap="Frequency of MHW Events in North America", fig.height=5}
ggMhwNamerFreq <- mhwYrTidy %>% 
  filter(region %in% namerRegs) %>% 
  ggplot() + 
  geom_histogram(aes(x=anomDays)) +
  theme_bw() + 
  labs(x="MHW Days per Year", y="Frequency") +
  facet_wrap(~region) + 
  NULL

ggMhwNamerFreq
```


```{r namer-mhw-anom, fig.cap="Anomaly of MHW Events in North America", fig.height=8}

ggMhwNamerAnom <- mhwTidy %>% 
  filter(region %in% namerRegs) %>% 
  ggplot(aes(x=date)) +
  geom_line(aes(y=anomIntC, color="Cumulative Mean Intensity")) +
  geom_line(aes(y=sstAnom, color="Single-Day Anomaly")) +
  scale_color_manual(name = "", values = c("Cumulative Mean Intensity" = "darkblue", "Single-Day Anomaly" = "darkgreen")) +
  facet_wrap(~region) +
  labs(x="Year",y="SST Anomaly") +
  theme_bw() + 
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  ) +
  NULL

ggMhwNamerAnom
```

### MHWs vs. CPUE

Let's look at annual cumulative mean intensity (sum of SST anomalies in the calendar year / number of MHW-days in the calendar year) vs. trends in aggregate abundance (total CPUE across all species) for the whole region. The two metrics of CPUE change used are proportion change CPUE vs. last year ((cpueThisYear - cpueLastYear)/cpueLastYear), and proportion change CPUE relative to the long-term mean ((cpueThisYear - cpueSeriesMean)/cpueSeriesMean). 

```{r namer-mhw-cpue-time, fig.cap="MHW Cumulative Mean Intensity and Change in Total CPUE Over Time", fig.height=8}
ggMhwNamerCpue <- namerSummary %>% 
  filter(year>=1982) %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=year)) + 
  geom_line(aes(y=anomIntC, color="Cumulative Mean MHW Intensity")) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  scale_color_manual(name = "", values = c("Cumulative Mean MHW Intensity" = "darkblue", 
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange")) +
  facet_wrap(~region) +
  labs(x="Year",y="SST Anomaly or Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL


ggMhwNamerCpue_save <- namerSummary %>% 
  filter(year>=1982) %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=year)) + 
  geom_line(aes(y=anomIntC, color="Cumulative Mean MHW Intensity")) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  scale_color_manual(name = "", values = c("Cumulative Mean MHW Intensity" = "darkblue", 
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange")) +
  facet_wrap(~region, ncol=4) +
  labs(x="Year",y="SST Anomaly or Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL
ggsave(ggMhwNamerCpue_save, filename=here("results","line_plots_MHW_CPUE_north_america.png"), dpi=160, width=8, height=4)

ggMhwNamerCpue
```

Rather than comparing both MHW intensity and CPUE trends over time, we can just plot cumulative mean intensity against the CPUE trend in that year:

```{r namer-mhw-vs-cpue, fig.cap="MHW Cumulative Mean Intensity vs. Change in Total CPUE", fig.height=8}
ggMhwNamerCpue3 <- namerSummary %>% 
  filter(year>=1982) %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=anomIntC, y=wtMtDiffProp)) + 
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~region) +
  labs(x="SST Anomaly",y="Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL

ggMhwNamerCpue3_save <- namerSummary %>% 
  filter(year>=1982) %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=anomIntC, y=wtMtDiffProp)) + 
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~region, ncol=4) +
  labs(x="SST Anomaly",y="Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL
ggsave(ggMhwNamerCpue3_save, filename=here("results","scatter_plots_MHW_CPUE_north_america.png"), dpi=160, width=8, height=4)


ggMhwNamerCpue3
```


```{r namer-mhw-5-spp, fig.cap="MHW Cumulative Mean Intensity vs. Single-Species Change in Total CPUE", fig.height=8, fig.scale=1.2}

ggMhwNamerCpue2 <- namerSppSummary %>% 
  mutate(sppReg = paste0(region, spp)) %>% 
  filter(sppReg %in% topspp$sppReg,
         wtMtDiffProp < 5) %>% # Keep only spp*region combos of interest
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=anomIntC, y=wtMtDiffProp)) + 
  geom_point() +
  geom_smooth(method="lm") +
  facet_wrap(~sppReg, ncol=5) +
  labs(x="SST Anomaly",y="Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom",
    strip.text = element_text(size = 5)
  ) +
  NULL

ggMhwNamerCpue2
```

### Evaluating a yearly MHW index

At Thomas's suggestion, let's try using a yearly definition of MHWs rather than a daily one, to match the time scale of the trawl data. In the histograms below I've imputed zeros for NA values (no MHW) just to see the data distribution.

```{r namer-mhw-yearly, fig.cap="Distributions of Yearly MHW Estimates Across Regions", fig.height=6, fig.scale=1.2}

ggMhwYrNamerFreq <- mhwYrTidy %>% 
  mutate_all(~replace(., is.na(.), 0)) %>% # replace all na values with zeros
  filter(region %in% namerRegs) %>% 
  ggplot() + 
  geom_histogram(aes(x=sstYrAnom)) +
  theme_bw() + 
  labs(x="Yearly Anomaly", y="Frequency") +
  facet_wrap(~region) + 
  NULL

ggMhwYrNamerFreq
```

This approach calculated exactly 3 MHW-years per region, I think because we took the 90th percentile of the same number of years: 

```{r namer-mhw-yearly-stats}
mhwYrTidy %>% 
  filter(!is.na(sstYrAnom)) %>% 
  group_by(region) %>% 
  summarise(mhw_years=n())
```

Let's see what happened to CPUE in those regions in those years; these are the same plots as above but now MHW years are indicated with an asterisk. In plots with fewer than three asterisks, there aren't trawl data available in the MHW year. 

```{r namer-mhw-yearly-cpue, fig.cap="Yearly MHWs and Change in Total CPUE Over Time", fig.height=8}

ggMhwYrNamerCpue <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=year)) + 
  geom_point(aes(y=sstYrAnom, color="Yearly MHW Index"), shape=8, size=2.5) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  scale_color_manual(name = "", values = c("Yearly MHW Index" = "darkblue", 
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange")) +
  facet_wrap(~region) +
  labs(x="Year",y="SST Anomaly or Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL

ggMhwYrNamerCpue_save <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  ggplot(aes(x=year)) + 
  geom_point(aes(y=sstYrAnom, color="Yearly MHW Index"), shape=8, size=2.5) +
  geom_line(aes(y=wtMtDiffProp, color="Prop. Change CPUE vs. Last Year")) + 
  geom_line(aes(y=wtMtAnomProp, color="Prop. Change CPUE vs. Mean")) +
  scale_color_manual(name = "", values = c("Yearly MHW Index" = "darkblue", 
                                           "Prop. Change CPUE vs. Last Year" = "darkred",
                                           "Prop. Change CPUE vs. Mean" = "darkorange")) +
  facet_wrap(~region, ncol=4) +
  labs(x="Year",y="SST Anomaly or Prop. Change CPUE") + 
  theme_bw() +
  theme(
    legend.position="bottom"
  ) +
  NULL

ggsave(ggMhwYrNamerCpue_save, filename=here('results','line_plots_MHW_yr_CPUE_north_america.png'), dpi=160, width=8, height=4)

ggMhwYrNamerCpue
```

How does catch vary in MHW years vs. non-MHW years?

```{r namer-mhw-yearly-boxplot1, fig.cap="CPUE Proportional Anomaly in MHW vs. non-MHW years", fig.height=8}

ggMhwYrNamerCpueBox1 <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  mutate(MHW = ifelse(!is.na(sstYrAnom), "yes","no")) %>% 
  ggplot() + 
  geom_boxplot(aes(x=MHW, y=wtMtAnomProp)) + 
  facet_wrap(~region) +
  theme_bw() +
  theme(
    legend.position="none"
  ) +
  NULL

ggMhwYrNamerCpueBox1
```

Here are the same plots using proportional difference from the last year instead of CPUE anomaly. 

```{r namer-mhw-yearly-boxplot2, fig.cap="CPUE Proportional Year-Over-Year Difference in MHW vs. non-MHW years", fig.height=8}

ggMhwYrNamerCpueBox2 <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  mutate(MHW = ifelse(!is.na(sstYrAnom), "yes","no")) %>% 
  ggplot() + 
  geom_boxplot(aes(x=MHW, y=wtMtDiffProp)) + 
  facet_wrap(~region) +
  theme_bw() +
  theme(
    legend.position="none"
  ) +
  NULL

ggMhwYrNamerCpueBox2_save <- namerSummary %>% 
  left_join (mhwYrTidy %>% filter(region %in% namerRegs), by=c("region","year")) %>% 
  mutate(MHW = ifelse(!is.na(sstYrAnom), "yes","no")) %>% 
  ggplot() + 
  geom_boxplot(aes(x=MHW, y=wtMtDiffProp)) + 
  facet_wrap(~region, ncol=4) +
  theme_bw() +
  theme(
    legend.position="none"
  ) +
  NULL
ggsave(ggMhwYrNamerCpueBox2_save, filename=here("results","boxplot_MHW_yr_CPUE_north_america.png"),dpi=160,width=8,height=4)

ggMhwYrNamerCpueBox2
```

Fitting a mixed-effects model to all the species-level data (not just top species) with random slopes and intercepts for species and region, `wtMtDiffPropAbs ~ sstYrAnom + (1 + sstYrAnom|spp) + (1 + sstYrAnom|region)` (note this is *absolute* change, so we are treating species that had increases or decreases in biomass as equivalent):

Note that the previous runs of this model had `NA` for years when there was no SST anomaly; replacing those values with 0 changed the results. 

```{r lme, results='asis'}
library(lme4)
library(broom.mixed)
library(kableExtra)

namerSppMhw <- namerSppSummary %>% 
  left_join(mhwYrTidy, by=c("year","region")) %>% 
  mutate(mhw_yes_no = ifelse(is.na(sstYrAnom),"no","yes"),
         sstYrAnom = replace_na(sstYrAnom, 0)) %>%  # after identifying MHW years, replace NAs with zeros
  filter(!is.na(wtMtDiffPropAbs),
         wtMtDiffPropAbs < Inf,
         wtMtDiffPropAbsLog < Inf,
         wtMtDiffPropAbsLog > -Inf) 

# LME pooled across all regions
lme1 <- lmer(wtMtDiffPropAbs ~ sstYrAnom + (1 + sstYrAnom|spp) + (1 + sstYrAnom|region), data = namerSppMhw) %>% tidy() # note that results are totally different if non-MHW years are classified as 0 rather than NA

kable(lme1) %>% kable_classic(html_font = "Cambria")


# for(i in unique(namerSppMhw$region)) {
#   lme1 <- lmer(wtMtDiffPropAbs ~ sstYrAnom + (1|spp), data=namerSppMhw[namerSppMhw$region==i,]) %>% 
#     tidy() 
#   
#   kable(lme1, caption=paste0(i)) %>% 
#     kable_classic(html_font = "Cambria")
# }

```

With the response variable on a log scale: `wtMtDiffPropAbsLog ~ sstYrAnom + (1 + sstYrAnom|spp) + (1 + sstYrAnom|region)`

```{r lme2, results='asis'}

# LME pooled across all regions
lme2 <- lmer(wtMtDiffPropAbsLog ~ sstYrAnom + (1 + sstYrAnom|spp) + (1 + sstYrAnom|region), data = namerSppMhw) %>% tidy() # note that results are totally different if non-MHW years are classified as 0 rather than NA

kable(lme2) %>% kable_classic(html_font = "Cambria")

```

Fitting a series of single-species models and plotting the results

```{r single-spp models}
library(purrr)

namerSppLm <- namerSppMhw %>% 
  group_by(spp, region) %>% 
  nest() %>% 
  mutate(
    model = map(data, ~lm(wtMtDiffProp ~ sstYrAnom, data = .x)), # not using absolute vals here on purpose -- curious about the sign of change 
    tidymodel = map(model, tidy) 
  ) %>% 
  unnest(tidymodel) %>% 
  filter(!term=="(Intercept)") %>%
  select(-data, -model) %>%
  ungroup()

namerSppLm %>% 
  arrange(desc(spp)) %>% 
  filter(estimate < 10000) %>% # get rid of ridiculous outliers
  mutate(spp = factor(spp, unique(spp)),
         signif = ifelse(p.value <= 0.05, "p <= 0.05","p > 0.05")) %>%
  ggplot(aes(x=spp, y=estimate, ymin=estimate-std.error, ymax=estimate+std.error, color=signif, fill=signif)) +
  geom_pointrange(size=0.7) +
  scale_color_manual(values=c('p <= 0.05'='#3A4ED0','p > 0.05'='grey85')) +
  geom_hline(yintercept=0, color="black") +
  labs(x=NULL, y="Proportional Biomass Change \n                 ") +
  #  scale_y_continuous(breaks=seq(-30, 20, 10), limits=c(-35, 20)) + 
  coord_flip() + # after this point x and y are reversed 
  facet_wrap(~region, ncol=length(unique(namerSppMhw$region)), scales = "free_x") + 
  theme_bw() +
  theme(legend.position = "none",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        #     text=element_text(family="sans",size=12,color="black"),
        #      axis.title=element_text(family="sans",size=14,color="black"),
        #      axis.text=element_text(family="sans",size=12,color="black"),
        axis.title.y = element_blank(),
        axis.text.y = element_blank()
        #      plot.margin=unit(c(5.5, 15, 5.5, 5.5), "points")
  ) +
  NULL

```

I already removed one major outlier, but it's still hard to see what's going on around the zero line. Let's zoom in: 

```{r single-spp models 2}

namerSppLm %>% 
  arrange(desc(spp)) %>% 
  filter(estimate < 10000) %>% # get rid of ridiculous outliers
  mutate(spp = factor(spp, unique(spp)),
         signif = ifelse(p.value <= 0.05, "p <= 0.05","p > 0.05")) %>%
  ggplot(aes(x=spp, y=estimate, ymin=estimate-std.error, ymax=estimate+std.error, color=signif, fill=signif)) +
  geom_pointrange(size=0.7) +
  scale_color_manual(values=c('p <= 0.05'='#3A4ED0','p > 0.05'='grey85')) +
  geom_hline(yintercept=0, color="black") +
  labs(x=NULL, y="Proportional Biomass Change \n                 ") +
  scale_y_continuous(limits=c(-50, 50), breaks=seq(-50, 50, 25)) + 
  coord_flip() + # after this point x and y are reversed 
  facet_wrap(~region, ncol=length(unique(namerSppMhw$region))) + 
  theme_bw() +
  theme(legend.position = "none",
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank()
  ) +
  NULL


```


